---
title: "Computing the LD-Corrected Gene-Level Response"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing the LD-Corrected Gene-Level Response}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

THREAD models a gene-level response derived from GWAS marginal effect estimates
and local linkage disequilibrium. For each gene, `compute_response()` returns:

- `final_y`, the LD-corrected response;
- `d0`, `d1`, and `d2`, the heteroscedastic variance coefficients;
- `n_snps`, the number of variants used;
- `status`, the computation status;
- optional LD completeness and PSD-repair diagnostics.

## Locate the installed example inputs

```{r locate-example}
library(THREAD)

example_dir <- system.file(
  "extdata",
  "example",
  package = "THREAD",
  mustWork = TRUE
)

matches_path <- file.path(
  example_dir,
  "example_gene_snp_matches.rds"
)

gwas_path <- file.path(
  example_dir,
  "example_gwas.tsv.gz"
)

example_X <- readRDS(
  file.path(example_dir, "example_x.rds")
)
```

The example LD file is stored as:

```text
example_ld_chr1.rds
```

The `{chr}` placeholder in `ld_file_template` allows
`compute_response()` to locate chromosome-specific files.

## Compute the response

```{r compute-response}
observed_response <- compute_response(
  gene_snp_matches = matches_path,
  gwas_clean = gwas_path,
  ld_dir = example_dir,
  ld_file_template = "example_ld_chr{chr}.rds",
  match_mode = "snp_name",
  genes = rownames(example_X),
  chromosomes = "1",
  n_cores = 1L,
  max_genes_per_chr = NULL,
  max_snps_per_gene = Inf,
  include_diagnostics = TRUE,
  drop_failed = FALSE,
  verbose = TRUE
)

table(observed_response$status, useNA = "ifany")
head(observed_response)
```

With `drop_failed = FALSE`, the returned object retains both successful and
unsuccessful records. Downstream model preparation uses records with
`status == "ok"`.

## Compare with the frozen example response

The installed example includes a response generated by the validated
implementation. It is used as a regression reference for package testing.

```{r compare-response}
expected_response <- readRDS(
  file.path(example_dir, "example_response.rds")
)

observed_response <- observed_response[
  order(observed_response$gene_name),
  ,
  drop = FALSE
]

expected_response <- expected_response[
  order(expected_response$gene_name),
  ,
  drop = FALSE
]

all.equal(
  observed_response[, c(
    "gene_name",
    "n_snps",
    "d0",
    "d1",
    "d2",
    "final_y",
    "status"
  )],
  expected_response[, c(
    "gene_name",
    "n_snps",
    "d0",
    "d1",
    "d2",
    "final_y",
    "status"
  )],
  tolerance = 1e-8,
  check.attributes = FALSE
)
```

## Variant matching modes

The recommended mode is:

```r
match_mode = "snp_name"
```

It requires consistent SNP identifiers across the GWAS table, the SNP-to-gene
mapping, and the LD reference.

The fallback mode is:

```r
match_mode = "snp_pos"
```

Position matching uses chromosome and position jointly. The GWAS summary
statistics, SNP-to-gene mapping, and LD reference must use the same genome
build.

## LD matrix reconstruction

Chromosome-level LD files may store an off-diagonal SNP pair in only one
direction. THREAD recovers the reciprocal entry and constructs a symmetric
gene-level LD matrix.

Pairs absent in both directions are treated as zero and their fraction is
retained as a diagnostic. When sparse pairwise storage produces an indefinite
symmetric matrix, negative eigenvalues are clipped to zero and the retained
spectrum is rescaled to preserve the trace.

## Full-size analyses

A complete analysis typically processes all autosomes and restricts response
calculation to genes present in the final expression matrix.

```{r full-analysis-template, eval=FALSE}
checkpoint_dir <- file.path(tempdir(), "THREAD_response_checkpoints")

response <- compute_response(
  gene_snp_matches = gene_snp_matches,
  gwas_clean = gwas_clean,
  ld_dir = ld_directory,
  match_mode = "snp_name",
  genes = rownames(X),
  chromosomes = 1:22,
  n_cores = 20L,
  max_genes_per_chr = NULL,
  max_snps_per_gene = Inf,
  output_dir = checkpoint_dir,
  save_by_chr = TRUE,
  resume = TRUE,
  include_diagnostics = TRUE,
  drop_failed = FALSE,
  verbose = TRUE
)
```

The installed example uses only a small chromosome 1 subset and must not be
interpreted as a complete disease analysis.
