---
title: "Getting Started with THREAD"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with THREAD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

THREAD integrates an LD-corrected gene-level GWAS response with a
gene-by-cellular-label expression matrix. It uses a heteroscedastic
Dirichlet process mixture to infer latent gene modules and module-specific
cellular coefficients.

The shortest workflow starts from:

1. a gene-by-cellular-label expression matrix, `X`;
2. a response table containing `gene_name`, `final_y`, `d0`, `d1`, and `d2`.

The installed example data are deliberately small. They are intended to
demonstrate the software interface and must not be used for biological
interpretation.

## Load the installed example data

```{r load-example}
library(THREAD)

example_dir <- system.file(
  "extdata",
  "example",
  package = "THREAD",
  mustWork = TRUE
)

example_X <- readRDS(
  file.path(example_dir, "example_x.rds")
)

example_response <- readRDS(
  file.path(example_dir, "example_response.rds")
)

dim(example_X)
head(example_response)
table(example_response$status, useNA = "ifany")
```

The rows of `example_X` are genes and the columns are cellular labels.
The response is retained on its original LD-corrected scale. Negative
`final_y` values, when present, are not truncated or log transformed.

## Prepare model input

`prepare_input()` aligns genes between the expression matrix and the
response, retains response records with `status == "ok"`, generates initial
cluster labels, and recommends prior hyperparameters when priors are not
supplied.

```{r prepare-input}
thread_data <- prepare_input(
  X = example_X,
  response = example_response,
  K_init = 3L,
  min_genes = 10L,
  seed = 2028L,
  verbose = TRUE
)

dim(thread_data$x_train)
table(thread_data$z_init)
thread_data$priors
```

Gene2vec initialization is optional. When no embedding is supplied, THREAD uses
a k-means fallback based on the aligned response and expression matrix.

## Fit the THREAD model

The settings below are intentionally short so that the vignette builds
quickly. A production analysis should use substantially longer chains and
should evaluate stability across seeds.

```{r fit-model}
fit <- run_thread(
  thread_data = thread_data,
  alpha = 0.1,
  m = 2L,
  n_iter = 60L,
  burnin = 30L,
  thin = 1L,
  seed = 2028L,
  save_scales = FALSE,
  verbose = TRUE
)

length(fit$samples)
summary(fit$diagnostics$cluster_counts)
```

## Obtain a point-estimate partition

THREAD summarizes posterior label draws through a posterior similarity matrix.
The point-estimate partition is selected by PAM silhouette within a local
window around the posterior-supported number of major modules.

```{r partition}
partition <- get_partition(
  fit = fit,
  min_cluster_size = 1L,
  mode_window = 1L,
  return_dissimilarity = FALSE,
  verbose = TRUE
)

partition$K
table(partition$assignments)
```

## Test module-cellular-label associations

`test_significance()` combines:

- feasible generalized least squares with BH FDR control;
- Bayesian local false sign rate;
- a dual-significance rule requiring both criteria.

```{r significance}
significance <- test_significance(
  fit = fit,
  partition = partition,
  thread_data = thread_data,
  fdr_thr = 0.05,
  lfsr_thr = 0.05,
  verbose = TRUE
)

head(significance$table)
significance$module_diagnostics
subset(significance$table, dual_significant)
```

The example contains only 24 genes but 31 cellular predictors. Consequently,
some or all example modules can be underdetermined for frequentist inference.
For such modules, THREAD retains descriptive coefficients but sets the
frequentist standard error, P value, and FDR to `NA`. They cannot be declared
dual-significant. This behavior is intentional and allows the example to test
the package interface without presenting the miniature dataset as a valid
biological analysis.

## Rank genes and cellular labels

```{r scores}
gene_results <- gene_score(
  partition = partition,
  coefficients = significance$coefficients,
  x_train = thread_data$x_train,
  top_n = 10L,
  verbose = TRUE
)

head(gene_results$top_genes)

cellular_results <- cell_type_score(significance)
head(cellular_results)
```

The gene score is an expression-weighted prioritization score rather than an
independent significance test. Biological conclusions should combine module
enrichment, transcription-factor analysis, external validation, and other
downstream evidence.
