---
title: "THREAD Input Formats and Workflow Entry Points"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{THREAD Input Formats and Workflow Entry Points}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Modular workflow

THREAD is organized as a modular workflow. Users may start from raw single-cell
data and raw GWAS summary statistics, or enter at a later stage when prepared
objects are already available.

```text
single-cell expression
    -> preprocess_scrna()
    -> gene-by-label expression matrix X

raw GWAS summary statistics
    -> preprocess_gwas()
    -> clean GWAS table
    -> match_snps_to_genes()
    -> SNP-to-gene mapping
    -> compute_response()
    -> gene-level response

X + response
    -> prepare_input()
    -> run_thread()
    -> get_partition()
    -> test_significance()
    -> gene_score() and cell_type_score()
```

## Expression input

The model expression matrix must contain:

- genes in rows;
- cellular labels in columns;
- unique, non-missing gene row names;
- finite numeric values.

```{r}
dim(X)
head(rownames(X))
head(colnames(X))
```

Users who already have a gene-by-label matrix can pass it directly to
`prepare_input()`.

For a Seurat object or a gene-by-cell matrix, use:

```{r}
X <- preprocess_scrna(
  object = seurat_object,
  subtype_col = "cell_subtype",
  already_normalized = FALSE
)
```

`preprocess_scrna()` is an aggregation helper. It does not replace upstream
single-cell quality control, doublet removal, batch correction, integration,
or cell annotation.

## GWAS summary statistics

The standard GWAS fields are:

```text
chrom
pos
rsid
beta
se
```

Optional fields can include MAF, INFO score, P value, and sample size.

```{r}
gwas_clean <- preprocess_gwas(
  sumstats = "gwas.tsv.gz",
  col_map = c(
    chrom = "CHR",
    pos = "BP",
    rsid = "SNP",
    beta = "BETA",
    se = "SE"
  )
)
```

## Gene annotation and SNP-to-gene mapping

A gene annotation table must provide a gene identifier, chromosome, start
position, and end position. THREAD maps SNPs to the annotated gene body extended
by a user-defined genomic window.

The standard mapping output contains:

```text
gene_name
snp_name
snp_pos
chrom
```

```{r}
gene_snp_matches <- match_snps_to_genes(
  gwas = gwas_clean,
  gene_annotation = gene_annotation,
  window = 10000
)
```

A SNP can map to more than one gene when gene windows overlap.

## LD reference files

Chromosome-level LD files are supplied through `ld_dir` and
`ld_file_template`. In SNP-name mode, the required fields are:

```text
SNP_A
SNP_B
R
```

In chromosome-position mode, the LD table must also contain the base-pair
positions used by the configured `BP_A` and `BP_B` columns.

GWAS, mapping, annotation, and LD resources must use compatible SNP identifiers
and genome builds.

## Gene-level response

A response table used by `prepare_input()` must contain:

```text
gene_name
final_y
d0
d1
d2
```

A table generated by `compute_response()` also contains `n_snps`, `status`,
and optional LD diagnostics.

```{r}
response <- compute_response(
  gene_snp_matches = gene_snp_matches,
  gwas_clean = gwas_clean,
  ld_dir = "path/to/ld",
  match_mode = "snp_name",
  genes = rownames(X),
  drop_failed = FALSE
)
```

Negative values of `final_y` are retained. They are not truncated or log
transformed.

## Prepared model input

```{r}
thread_data <- prepare_input(
  X = X,
  response = response,
  gene2vec = gene2vec,
  K_init = 10L,
  seed = 2028L
)
```

`prepare_input()` aligns gene identifiers, uses response records with
`status == "ok"`, initializes gene labels, and recommends prior
hyperparameters when required.

Gene2vec is optional. When it is absent, THREAD uses a k-means fallback.

## Model fitting

```{r}
fit <- run_thread(
  thread_data = thread_data,
  alpha = 0.1,
  m = 3L,
  n_iter = 30000L,
  burnin = 25000L,
  thin = 1L,
  seed = 2028L
)
```

Production analyses should evaluate posterior stability, module-count
trajectories, and robustness across seeds.

## Point-estimate partition

```{r}
partition <- get_partition(
  fit = fit,
  min_cluster_size = 10L,
  mode_window = 3L
)
```

## Association inference

```{r}
significance <- test_significance(
  fit = fit,
  partition = partition,
  thread_data = thread_data,
  fdr_thr = 0.05,
  lfsr_thr = 0.05
)
```

An association is dual-significant only when it satisfies both the frequentist
FDR criterion and the Bayesian local false sign rate criterion.

For modules with no more genes than predictors, descriptive coefficients are
retained, but frequentist standard errors, P values, and FDR values are set to
`NA`.

## Downstream scores

```{r}
gene_results <- gene_score(
  partition = partition,
  coefficients = significance$coefficients,
  x_train = thread_data$x_train
)

cellular_results <- cell_type_score(significance)
```

These scores are downstream prioritization summaries rather than independent
significance tests.

## Supported entry points

| Available inputs | First required THREAD function |
|---|---|
| Raw single-cell data and raw GWAS | `preprocess_scrna()` and `preprocess_gwas()` |
| Gene-by-label matrix and raw GWAS | `preprocess_gwas()` |
| Clean GWAS and gene annotation | `match_snps_to_genes()` |
| SNP-gene map, clean GWAS, and LD | `compute_response()` |
| Expression matrix and response | `prepare_input()` |
| Prepared `thread_data` | `run_thread()` |
| Fitted `thread_fit` | `get_partition()` |
| Fit, partition, and model data | `test_significance()` |
| Significance output | `gene_score()` and `cell_type_score()` |

This modular structure allows externally prepared inputs to enter the workflow
without forcing users to rerun earlier stages.
